
# A brief R programming crash course
To get started with making our own evolutionary model, we need to make a simple program in R that tracks an evolving population. 

To this end, you need to know something about the ingredients that simple programs use, namely variables, vectors/sequences and for-loops. Some of this may have already featured in your statistics or introduction to data science modules, but at this stage, getting some repetition about the sometimes bamboozling inner workings of R certainly does not hurt. Even the most field-oriented among you will use R extensively during your research projects, so the more exposure you can get to this environment, the better.

Try to type along with the examples below and see whether they work. Give me a shout if something looks odd. 

## Variables
A variable provides us with a named storage of data, allowing you to retrieve that bit of data when you want it. A variable in R can contain any type of data (e.g., a single number, a string of text, a vector of multiple values, etc). 

We assign data to a variable using a `=` or `<-` operator (both can be used interchangeably) For example:
```{r}
# create a variable named var.1 and assign a value to it using the <- operator
var.1 <- 25000

# create a variable named var.2 and assign a vector, c(), of numbers to it using the = operator
var.2 = c(0,1,9.5,200)

# a bit of text, surrounded by single '' or double "" quotes
some.text <- "Swanpool beach"
```

### Naming variables
Variables can have any name, but cannot start with a number and can only contain dot `.` and underscore `_` characters as special characters.

```{r, error=TRUE,echo=TRUE, collapse=T}
# this fails
2times_song_release_year <- 1999
```

```{r}
# this is fine
release_year_2times <- 1999
```

```{r, error=TRUE,echo=TRUE,collapse=T}
# this fails too
release%year <- 1999
```


### Changing the value stored by variables
You can easily change values of variables, simply by assigning them a new one! The new value can be of a completely different type.     
```{r,collapse=T}
# print the value of a variable, showing its current value
print(var.1)

# change the value
var.1 <- "some text instead!"

# print the new value
print(var.1)
```

You can always check the type of your variable using the `typeof()` function:
```{r, collapse=TRUE}
typeof(var.1)
typeof(var.2)
```
As you see, `typeof` does not tell you `var.2` is a vector of multiple values, but just focuses on the type of the values contained in them (in this case a `double`, which is another word for floating point value. If you want to find out if something is a vector of values, you have to look at the length of the variable (see below).

## Vectors and element selection
As you have seen in the example of `var.2` above, variables can point to collections of multiple values (vectors). To create a vector, we use the `c()` function. We can subsequently get values of (groups of) individual elements by using the `[[]]` operator:

```{r,collapse=T}
# 1. create a vector of character elements using c()
animals <- c("Opossum","Dog","Pallas's leaf warbler")

# 2. select a single element using the double square brackets [[]]
print(animals[[3]])


# 3. obtain a selection of 
# multiple values using [] and
# a vector of elements you want to select
print(animals[c(2,3)])


# 4. assign a new value to an existing element
animals[[1]] <- "Click beetle"
print(animals)
```

### Exercises
 * What happens to the vector `animals` if you assign the name of your favorite animal (in quotes `""`, because text) to element `40`?
 ```{r,echo=F}
 animals[[40]] = "Blue-cheeked bee eater"
 ```
 
 * Assign the number 0.5 to element `10` of the `animals` vector. Try to multiple this value by 10, by typing
 
 ```{r,eval=F}
 animals[[10]]*10
 ```
what happens? Use `typeof(animals[[10]])` to see what is going on.

## Information about your vector
You can obtain information about your vector by using a range of different functions, such as 

 * `sum()`: sums all values (only works with numbers obvz)  
 * `unique()` lists all unique values  
 * `sort()` sorts values. 
and there are lots more. 
 
 
 Most often, however, you will use the `length()` function to obtain its size:
```{r, collapse=T}
length(x=animals)
```

### Exercise
Using the standard `iris` dataset (available within in `R` by typing `iris` on the command line), find out the number of unique values for the column `Sepal.Width`. If you see `iris` but don't know how you can access the column, try to search for it on the web.

## Sequences
Vectors are also used to contain sequences of numbers, which can either be generated with the from-to operator `:` or with the more explicit `seq()` function:
```{r,collapse=T}
# sequence from 1 to 10
some.seq <- 1:10

# sequence from 10 to 1
reverse.seq <- 10:1

# same, but using the sequence in some.seq and the rev() 'reverse' function:
reverse.seq.2 <- rev(some.seq)

# sequence from 10 to 1, using the seq() function, with a negative increment
another.seq <- seq(from=10,to=1,by=-1)

# the seq() function is great if you want to take bigger steps:
steps.of.5 <- seq(from=3,to=29,by=5)
```

We can also make vectors using repetitions of values, through `rep()`:
```{r}
# a vector of 300 zeros
only.zeros <- rep(x=0,times=300)

# a vector of 0,1 values, repeated until a length of 13 is achieved
zero.one <- rep(x=c(0,1),length.out=13)
```

## The important bit: `for` loops
The point about programming is that you want to repeat actions multiple times. To this end, R offers you two different looping constructs: the `while()` loop and the `for` loop. 

In this practical, we focus on the `for` loop as it is more commonly used than `while()`. Best to begin by an example:
```{r, collapse=T}
# you need some vector of something
# otherwise there will
# be nothing to loop over
# Here, I use a vector of numbers from 5 up to (& including) 10
some.vec <- 5:10

# component 1: the for statement within 
# parentheses ()
for (iterator in some.vec) 
{
    # component 2: the for body within curly braces {} 
    # in which tasks 
    # are performed
    # repeatedly for each value of iterator
    print(iterator)
    
    # I am also printing something else
    # for the sake of illustration
    print("Some text.")
}
```

From the printed output, you see that the `iterator` gets a new value every time the loop goes round. The message `"Some text"` is also repeatedly printed, but this value does not change.

### Dissecting the for loop
The loop starts with the keyword `for`, followed by parentheses `()`. Within those parentheses, we always start by declaring a new variable, which -- in this case -- I have called `iterator`, but you could have given it *any* valid name, like `index` or `boring.practical`. 

Each time the `for` loop goes round, the `iterator` variable gets assigned a new value. Which value? An element from a vector, which we have called here `some.vec`. The `in` keyword in between `iterator` and `some.vec` performs the assigning of these consecutive elements from `some.vec` to `iterator`. 

Hence, the first time the loop goes round, `iterator` receives the value of `5`, as this is the first element of `some.vec`. Subsequently, the `for` loop enters the bit in between curly braces `{}`. The code in this part will be executed every time `iterator` gets a new value. In this case, the code is `print(iterator)`, meaning it prints the *current* value of the `iterator` variable (which is `5`) is printed to the screen. After this the subsequent `print()` statement is executed, which prints `"Some text"` to the screen.

Next, the loop goes round again. Now the value of `iterator` will be 6. Again, `6` is printed by the `print()` statements in the `{}`-part and this goes on until all elements from `some.vec` have been assigned to `iterator`. Hence, the last value printed will be `10`, after which the `for` loop exits.

### Another example
Just to throw in another example that is slightly different, let us calculate the product of elements of two vectors and sum those products:
```{r,collapse=T}
# generate two vectors with numbers; 
# vectors have the same length
vec.1 <- c(0.3,0.9,1.2,0.1,3.5)
vec.2 <- c(5,8,12,3,7)

# check whether the length is indeed the same
stopifnot(length(vec.1) == length(vec.2))

sum <- 0

# loop not over the elements of the list
# but over the element indices
# (we can do so by creating a list from 
# 1 to the length of one of the vectors,
# i.e., 1,2,3,...)
for (idx in 1:length(vec.1))
{
    # use idx to obtain elements of both vectors
    # and add them to the total
    sum <- sum + 
        vec.1[[idx]] * vec.2[[idx]]
}
print(sum)
```

### Exercises
  * Can you think of why we don't we write ``for (idx in vec.1)`` as in the previous example?
  * Could you change the code to store the sum that you obtain
  in each iteration in a list? I.e., that list should have the values
```{r,echo=F}
vec.1 <- c(0.3,0.9,1.2,0.1,3.5)
vec.2 <- c(5,8,12,3,7)

# check whether the length is indeed the same
stopifnot(length(vec.1) == length(vec.2))

sum <- 0

cumul.list <- rep(0,times=length(vec.1))

# loop not over the elements of the list
# but over the element indices
for (idx in 1:length(vec.1))
{
    # use idx to obtain elements of both vectors
    # and add them to the total
    sum <- sum + 
        vec.1[[idx]] * vec.2[[idx]]
    
    cumul.list[[idx]] <- sum
}
```
```{r}
print(cumul.list)
```

OK now that we have some tools under our belts, let's get going with evolution!!

