# An evolutionary model of the hawk-dove game

Why do we often find that competitors 
exhibit ritualized displays
(e.g., threats such growling, baring one's teeth)
to decide fights with competitors,
rather than engaging in escalated fights in order to kill the opponent?

This question has been addressed by a simple evolutionary model, 
called the Hawk-Dove game. This influential model aims to sketch out how 
aggressive versus more restrained behaviours evolve 
when individuals compete with each other over resources. 

Here we analyze a deterministic version of this
model, by focusing on a population consisting of i) aggressive hawks (denoted by $H$) who
always fight and do not retreat; and ii) timid Doves (denoted by $D$), who may perform some
threat display, but eventually will always retreat if the other individual
starts a fight. 

## Payoff matrices
We can summarize these interactions in a so-called payoff matrix, where the rows
reflect the payoff to a focal player who either plays strategy $H$ or $D$, when it encounters other $H$s or $D$s (columns).

| focal $\downarrow$  opponent $\rightarrow$       |   $H$   |  $D$     |
|---------------------|----------|----------|
|  $H$              |  $\frac{v-c}{2}$  | $v$ |
|  $D$              |  $0$  | $\frac{v}{2}$ |

Table: (\#tab:payoffHD) The payoff matrix of the Hawk-Dove game.

## Exercise
Can you explain in your words what each of the payoffs mean? Try to search for descriptions of the Hawk Dove game online if unclear. We will discuss the answer in class.

## A haploid population genetical model of the Hawk-Dove game
Although the Hawk-Dove game is often used as an example of evolutionary game theory, here we start from the ground up by developing a more rigorous model that is based on population genetics. Population genetics deals with genetic detail, so is more complete than evolutionary game theory. However, analyzing the formulas of the population genetics model is slightly more challenging, so this is why we will use R to simulate the model instead. 

By comparing whether predictions from evolutionary game theory match our population genetics model, we can see whether genetic detail matters, if at all. If predictions do not match, this may tell us that lack genetic assumptions in evolutionary game theory can lead to erroneous conclusions. Such exercises are worthwhile if you want to predict when a phenotypic gambit may be an accurate vs inaccurate description of the evolution of behaviour.

## Lots of crazy assumptions
For the sake of simplicity, we assume that being a hawk or dove is the result
of **just two alleles at a single haploid locus**! Moreover, reproduction occurs asexually. The population sizes are assumed to be infinite.
This is crazy of course, hardly any animals are asexual, let alone haploid. Moreover, even
fewer are chiefly the result of variation at a single locus only. 

The question is whether we would learn a similar amount from a hugely convoluted model in which we track the actual population size, have 1000s of loci or deal with sexual reproduction and all the resulting products of recombination.

## Allele frequencies
Let $p_{t}$ be the frequency of the allele coding for $H$ in the population at generation $t$, while $1-p_{t}$ is the frequency of the $D$ allele in the population at generation $t$. We assume that strategies are _pure_, meaning that an individual having the $H$ or $D$ allele will always be a hawk or dove respectively. Later on, we will consider a case where strategies are _mixed_, so that $p_{t}$ does not reflect the frequency of the $H$ allele, but the probability with which each individual plays $H$ at each interaction.

## Life cycle
We assume that generations are discrete, implying that all individuals die after reproduction. Upon birth, a newborn individual fights or displays with other individuals a single time. From this interaction, it will collect a payoff as given in the payoff matrix above. The payoff is then proportional to the number of offspring the individual produces.

## Fitness
An individual reproduces according to the payoffs it has collected. Let us denote the fitness of a hawk and dove as $W_{H}$ and $W_{D}$ respectively, which we can interpret as the number of offspring produced by each of these types.

Starting with the fitness of a focal dove $W_{D,t}$ at generation $t$, we have
\begin{align}
W_{D,t} &= w_{0} + p_{t} (0) + \left(1-p_{t} \right) \frac{v}{2}
\end{align}

 * the first element on the right-hand side of the equation above reflects so-called 'baseline fitness', which is the fitness that this individual accrues during other activities than displaying. It is assumed that baseline fitness is similar for hawks and doves.
 * the second part reflects the probability that a dove meets a hawk (which has frequency $p_{t}$). It will then immediately retreat, hence collecting no payoff (but also incurring no costs), reflected by the $(0)$. 
 * the third part reflects the probability that a dove meets another dove (which has frequency $1-p_{t}$). In that case they will split the value of the resource, hence $v/2$.

## Exercise

1. Based on similar reasoning as for $W_{D,t}$ above, can you write down the fitness expression $W_{H,t}$ for a focal hawk in generation $t$? Again, assume that there is some baseline fitness $w_{0}$ accrued from other activities than fighting:

\begin{align}
W_{H,t} &= w_{0} + \ldots + \ldots
\end{align}

2. Can fitnesses $W_{H,t}$ and $W_{D,t}$ be negative? 

## Evolutionary dynamics
Now that we have derived fitness expressions of the dove and hawk alleles, let us consider how the frequencies of these alleles change over time. The frequency $p_{t+1}$ of the hawk allele at time $t+1$ is given by a surprisingly simple equation:

\begin{align}
p_{t+1} &= p_{t}  \frac{W_{H}}{\bar{W}_{t}}
\end{align}

This is a so-called recursion equation, tracking an allele frequency *recursively* from one generation to the next. Keep an eye on this equation, as we will need it later on. The term $\bar{W}_{t}$ in the expression above is the average fitness in the population in generation $t$, given by
\begin{align}
\bar{W} &= p_{t} W_{H} + \left(1 - p_{t} \right ) W_{D}
\end{align}

The expression for $p_{t+1}$ above tells you that if hawks do better than the average ($W_{H} > \bar{W}$), the frequency of the hawk allele will increase from one generation to the next, as $W_{H}/\bar{W} > 1$. Similarly, if hawks do worse than average, we have $W_{H}/\bar{W} < 1$ and the frequency of the hawk allele decreases.

One advantage of our super-simple model is that we do not have derive a separate expression for the change in frequency of the dove allele!! This is simply $1-p_{t+1}$. However, if we would consider more than two alleles per locus (or multiple loci), matters would be (far) more difficult.

## Relation to $\Delta p$ in lecture slides (only if interested)
In the lecture slides we have seen that the change in allele frequency was given by
\begin{align}
\Delta p = p_{t+1} - p_{t} = p_{t} \left ( 1 - p_{t} \right ) \frac{ W_{H} - W_{D}}{\bar{W}}
\end{align}

If you want to know how the above relates to the equation we have seen in the lecture slides, read this! If not interested, please skip ahead to the next section.

How do we get from our recursion equation $p_{t+1} = \ldots $ to the one above? 
The clue is in the $\Delta p = p_{t+1} - p_{t}$. If we substitute here for our recursion equation of $p_{t+1}$, we get:

\begin{align}
\Delta p &= p_{t+1} - p_{t} \\
&= p_{t}  \frac{W_{H}}{\bar{W}_{t}} - p_{t} \quad \text{substitute for }p_{t+1} \\
&= p_{t}  \frac{W_{H}}{\bar{W}_{t}} - p_{t} \frac{\bar{W_{t}}}{\bar{W}_{t}} \quad \text{common demoninator} \\
&= p_{t}  \frac{W_{H}}{\bar{W}_{t}} - p_{t} \frac{p_{t} W_{H} + \left (1  - p_{t}\right ) W_{D} }{\bar{W}_{t}} \quad \text{expanding }\bar{W} \\
&= p_{t} \left (1 - p_{t} \right )  \frac{W_{H}}{\bar{W}_{t}} - p_{t} \frac{\left (1  - p_{t}\right ) W_{D} }{\bar{W}_{t}} \quad \text{clubbing terms of }W_{H} \\
&= p_{t} \left (1 - p_{t} \right ) \frac{W_{H} - W_{D}}{\bar{W}_{t}} \quad \text{clubbing terms of }p_{t} \left(1-p_{t} \right) \\
\end{align}
and we are done. Quite a bit of algebra to show that $p_{t+1}$ and $\Delta p$ are just two sides of the same coin, the first one is a *recursion* equation, the second one a *difference* equation.

## Implementing an evolutionary algorithm in R
In order to asses whether hawks or doves win out, we need to iterate the population genetics recursion equation $p_{t+1} = p_{t} W_{H} / \bar{W}$ for a number of timesteps until there is no further change.

The `for` loop  that we briefly encountered is an ideal tool to do this. To get you started, I provide a brief part of an R-script - containing mistakes and omissions - that you can modify so that it can update the recursion equation over multiple time steps until evolution stops.

```{r, code=xfun::read_utf8("hawk_dove_exercise.r"),eval=F}
```
## Exercise
Can you improve the code above to obtain the change in frequency of the hawk allele over `max_time` timesteps? If successful, try to explore the parameter values $v$ and $c$.

## Exercise
Can you try to run your script multiple times, each time for a different initial value of the frequency of hawks? Does it matter where you start? Can you plot the output from your simulation either in $R$ or in Excel?
Plotting in R can be done with
```{r,echo=F}
v <- 2.0
c <- 4.0

w0 <- 5

max_time <- 100
p <- rep(NA,times=max_time)

p[[1]] <- 0.1

for (iter in 1:(max_time-1))
{
    wD <- p[[iter]] * 0 + (1-p[[iter]]) * v/2
    wH <- p[[iter]] * (v-c)/2 + (1 - p[[iter]]) * v
    
    wBar <- p[[iter]] * wH + (1-p[[iter]]) * wD
    
    
    p[[iter + 1]] <- p[[iter]] * wH / wBar
    
}
```

```{r,eval=T}
plot(x=1:max_time
     ,y=p
     ,ylim=c(0,1.0)
     ,xlab="generation"
     ,ylab="frequency Hawk allele")
```

## Exercise
From evolutionary game theory, we know that the equilibrium frequency $\hat{p}$ of hawks (i.e., the frequency at which there is no further change in the frequency of hawks) is $\hat{p} = v/c$. Do you get the same answer here? What would you conclude about the importance of genetic detail?



