

# Individual-based simulations and drift

Previously, the model of the        Hawk-Dove game that we have used was **deterministic**, where multiple runs for the same values of the parameters $v$ and $c$ always result in the same outcome. 

In reality, however, evolution does not work like that, as quite a number of biological processes are the result of **chance**. To give some examples: chance plays a big role in determining whether a mutation occurs or not, whether or not you happen to encounter a hawk rather than a dove, which gamete actually contributes to the next generation or which offspring manages to survive in order to reproduce. 

Introducing chance in models of evolution makes these models **stochastic**, meaning that any two runs of the same model (with the same values of the parameters) does not necessarily result in the same outcome.  Stochasticity in evolution introduces a phenomenon called 'genetic drift' which we will explore in this practical. Genetic drift can -- particularly in small populations -- dramatically affect the course of evolution!

## Before you start this practical
First, please be reminded to use one of the [Rstudio servers](https://rstudio01.cles.ex.ac.uk) rather than your local installation of R. This is because we will be using some custom-built R packages, which themselves require extensions like [Rtools](https://cran.r-project.org/bin/windows/Rtools/) that can be a bit of a pain to install on some computers. Luckily, this should all have been done for you on the Rstudio server.

Next, when logged in on the Rstudio server, please use R version `4.1.1.`

And finally, it may be worthwhile to remove variables from our previous session, which we do by typing in the following
```{r, eval=F, message=F}
# clean all R variables
rm(list=ls())
```

## Installing the `driftSim` package
Now, let's get started. In this part of the practical, we will explore a model that includes stochasticity. Such a model is provided in the R package `driftSim`. To install this package, do the following:

```{r, eval=F, message=F}
if (!require("devtools")) install.packages("devtools")

library("devtools") 
devtools::install_github("bramkuijper/driftSim")

```
If you get a message like `Skipping install of 'driftSim' from a github remote, the SHA1 (27e9c52c) has not changed since last install. Use `force = TRUE` to force installation`, R is simply telling you the package has already been installed. Consequently, you can just move on to the next step.


Then we can load the library using:
```{r, eval=T, messages=F}
library("driftSim")
```

## Individual-based models
This package contains what we call an *individual-based simulation*, which is nothing more than a computer program that models individuals and their actions (e.g., surviving, fighting, reproducing). By contrast, in the previous set of exercises, we did not focus on individuals too much, but rather on an *analytical model* which tracked the population-wide frequency $p_{t}$ of hawks. Indeed, keeping track of a single population-wide frequency $p_{t}$ is a lot simpler than keeping track of thousands or even millions of individuals. However, such simplicity also comes also at a price, which will be the focus of this chapter.

While the simulation package `driftSim` runs within R, under the hood it is coded in the low-level programming language [C++](https://cplusplus.com/doc/tutorial/). This is because large simulations tend to be very slow in R, but run much faster in C++. Below you see a little snippet from this package, reflecting the properties of single individual:

```{c, eval=F}
struct Individual {
    bool is_hawk; // TRUE/FALSE individual is hawk or dove
    double payoff; // the value of the payoff
    double prob_hawk; // the probability that this individual develops as a hawk in early life (only used when strategies are mixed)
};
```

The program then simulates a **finite** population of $N$ of these individuals. You might remember that the analytical model in the previous chapter assumed infinitely many individuals, so while the individual-based model is an absolute loser when it comes to speed, it is already a winner when it comes to increased realism!

## Life cycle
At the start of each generation, each individual interacts with a randomly chosen other individual and either attacks (hawk) or displays (dove). It gets its payoffs based on the same payoff matrix as we used in the deterministic model in the previous chapter. Subsequently, we then produce $N$ newborn offspring from this population. Parents of each of these offspring are sampled from the previous generation, with individuals who have a larger payoff being more likely to be selected than individuals which have a lower payoff.

## Exercise: limiting assumptions
Individual-based simulations are often used to relax many of the limiting assumptions present in deterministic models, as writing computer code allows for much more flexibility than when using mathematical formulas. We have already seen how individual-based simulations have relaxed one assumption of the popoulation genetics mode, namely the fact that population sizes can now be finite, rather than infinite. 

Can you list three other limiting assumptions from the deterministic population genetics model that you would like to relax? For each assumption, can you think how relaxing the assumption could potentially affect results? (It is often much easier to spell out what is wrong with a model than to spell out how addressing it will change outcomes!)

## Running the `driftSim` package
The `driftSim` package contains a single function, called `runSimulation()`. Before running this function, please inspect the documentation by typing `?runSimulation()` in your R console, after which a help page should show up. 

Let us start with a simple example run with a population size of $N=10$, $v=1.0$, $c=2.0$ and a bunch of different parameters. Most are not so important, except that the initial frequency $p_{t=0}$ of hawks which is set at $p_{t=0} = 0.25$.
```{r,eval=T,collapse=T}
runSimulation(N=10,v=1.0,c=2.0,is_pure=T,mu=0,max_time=10,pHawk_init=0.25,output_nth_generation=1,sd_pHawkMixed=0)
```

In this way, however, we cannot use the data from the `runSimulation()` function, as it will simply be printed out to the console, but it is not stored anywhere. To make sure that the resulting data will be contained in a `data.frame`, we assign the return values from the `runSimulation()` function to a variable, like this:
```{r,eval=T,collapse=T}
my.data <- runSimulation(N=10,v=1.0,c=2.0,is_pure=T,mu=0,max_time=10,pHawk_init=0.25,output_nth_generation=1,sd_pHawkMixed=0)
```

The `my.data` variable points to a `data.frame` with the following columns:
 1. `generation`: generation time point
 2. `freq_Hawk`: the frequency of hawks in the population, the variable of interest
 3. `mean_pHawkMixed`: in case hawks and doves are the result of a mixed strategy, this is the (evolving) probability that any individual will develop as hawk at the start of its life. This is 0 in case `is_pure=T`, when we consider pure strategies of hawks and Doves
 4. `sd_pHawkMixed`: variation among individuals in their ability to develop as hawks (only nonzero when `is_pure=F`)
 
## Exercise: plotting the output from the `driftSim` package
Make a plot where plot `generation` on the $x$-axis and `freq_Hawk` on the $y$-axis. Re-run your simulation at least 10 times. What happens (typically) with the hawk allele? 

## Exercise: change the initial frequency of hawks
Now, change the initial frequency of hawks by changing the value of `pHawk_init` to `pHawk_init=0.75`. Run the simulation for a long time, for example `max_time=500`. What happens now with the frequency of hawks? What does that tell you about the differences between the analytical model and the current one? Do we still reach the predicted equilibrium frequency of hawks $\hat{p} = v/c$?

## Exercise: increase the population size
Now, try the same with a population size of $N=500$. What happens now?


